\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \thanks{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-01-27 move_schedules.tex
Cleaning section 3

 \begin{enumerate}[labelsep=3pt, topsep=2pt,itemsep=-0.25ex,parsep=1.0ex]

@@

Property M2 implies the decomposition is not fully maintained during the execution of \textit{RestoreRight} and \textit{RestoreLeft} below, which starkly contrasts the vineyards algorithm. 
In this way, we interpret move operations as making a tradeoff in granularity: whereas a sequence of adjacent transpositions $(i, i{+}1), (i{+}1, i{+}2), \dots, (j{-}1, j)$ generates $\lvert i - j \rvert $ valid decompositions in vineyards, an equivalent move operation $\mathrm{Move}(i,j)$ generates only one.
Indeed, Property M3 directly follows from this fact, as one may simply move each simplex $\sigma \in K$ into its order in $K'$ via insertion sort. Note that the number of valid decomposition produced by vineyards is bounded above by $O(m^2)$, since each pair of simplices $\sigma_i, \sigma_j \in K$ switches its relative ordering at most once is the homotopy assumptions from~\ref{sec:vineyards} hold.

@@

%\subsection{Proposed approach}
Let us begin with a brief overview of the pipeline, outlined in Algorithm~\ref{alg:schedule} below.
As before, we assume as input  a discrete $1$-parameter family of filtrations $\mathcal{K} = (K_1, K_2, \dots, K_n)$ of an abstract simplicial complex $K$ with $\lvert K \rvert =m$, and the goal being to compute their persistence diagrams.
Fix reindexing  bijections $f_i : K_i \to K_{i+1}$. 
Each $f_i$ induces a bijection $f_i^\ast : [m] \to [m]$, or equivalently, a permutation of the index set $[m]$. For each pair of filtrations $(K_i, K_{i+1})$, we assign the simplices of $K_i$ labels from index set $[m]$ in filtration order, relabeling $K_{i+1}$ accordingly using $f_i$. 
As $p$ is now a strictly increasing sequence, we may compute the longest increasing subsequence $\mathrm{LIS}(q)$ and use this subsequence to recover a longest common subsequence (LCS) between $(K_i, K_{i+1})$, which we denote later with $\mathrm{LCS}(K_i, K_{i+1})$. We pass $q$ and $\mathrm{LIS}(q)$ to our greedy scheduling algorithm, which returns as output an ordered set of move permutations  $\mathcal{S}$ of minimum size, which we call a \emph{schedule}.
Note one need not explicitly store the entire family of filtrations nor the schedules between them---Algorithm~\ref{alg:schedule} may be easily modified to be completely \emph{online}, keeping at most two filtrations and one decomposition in memory at any given time. 
% Not all valid schedules are created equal: some incur a much larger number of column operations, even if they have a minimal number of moves; we address this issue in Section~\ref{sec:schedule_cost}.
% Given the set of schedules for each pair $(K_i, K_{i+1})$, we proceed to computing an $R=DV$ decomposition for the first filtration $K_1$, and then execute the moves stored in $\mathcal{S}$ in sequence. During execution, we obtain a valid decomposition $R = DV$ for each filtration $K_1, K_2, \dots, K_n$ from which we extract persistence diagrams along the way.

@@

\subsubsection{The Vineyards Setting}\label{sec:continuous_setting}
In the vineyards (i.e., continuous time) case, we are given a homotopy $F : K \times [0,1] \to \mathbb{R}$ interpolating between  a pair of filtrations $K_0, K_1$ of $K$, with simplices  ordered according to $f_0 = F(\cdot, 0)$ and $f_1 = F(\cdot, 1)$, respectively. 
The choice of homotopy $F$ completely determines the number of adjacent transpositions that transform $K_0$ into $K_1$ via $F$, so to get a meaningful bound we must know something about $F$. 
If we assume that the curves $\tau \mapsto F(\cdot, \tau)$  are in general position in $[0,1]\times \mathbb{R}$, and that each pair of curves cross  at most once, then the set of adjacent transpositions $S_F$ that $F$ decomposes into is given by the pairs of simplices in $K$ with changes in their relative ordering:
\begin{align*}\label{eq:sf_schedule}
	S_F &= \{ (i,i+1) \, \mid \, f_0(\sigma_i) < f_0(\sigma_{i+1}), \;  f_1(\sigma_i) > f_1(\sigma_{i+1}), \; i \in [m] \} \\
	&= \{ s_1, s_2, \dots, s_k \} 
\end{align*}
%We assume, without loss of generality, that the underlying complexes $K_0$ and $K_1$ have the same cardinality. 
%Since we deal exclusively with essential filtrations, we may think of $K_1$ as simply a reordering of $K_0$ (or vice versa). 
Let $p = [m]$ and let $q = \mathrm{Im}\,f^\ast$, where $f^\ast : [m] \to [m]$ is the one-to-one correspondence induced by the reindexing bijection $f: K_0 \to K_1$. Then $S_F$ is given by the inversions between $p$ and $q$:
\[S_F = 
\mathrm{Inv}(p, q) = \{ \, (i,i+1) \, \mid \, p(i) < p(i+1), \; q(i) > q(i+1), \; i \in [m]\, \}
\]
The cardinality of $  S_F $ is exactly the \emph{Kendall}-$\tau$ distance~\cite{diaconis1977spearman} between $p$ and $q$: 
\begin{equation}\label{eq:kendall_dist}
	K_\tau(p, q) = \lvert \, \mathrm{Inv}(p, q) \, \rvert
\end{equation}
Geometrically, our assumptions on $F$ imply that the  functions $\tau\mapsto F(\cdot, \tau)$ belong to a class of $x$-monotone curves called \emph{pseudo-segments}.
%, i.e. parameterized curves whose behavior with respect to a certain restricted set of geometric predicates is invariant. 
This family includes 
the straight-line homotopy $F(\sigma, \tau) = (1 - \tau) f_0(\sigma) + \tau f_1(\sigma)$,   studied in the original vineyards paper~\cite{cohen2006vines}. Detecting all $k$ intersections of $m$ pseudo-segments is a well-studied problem in computational geometry that can be optimally solved in output-sensitive $O(m \log m + k)$ time by several algorithms~\cite{boissonnat2000efficient}, where $k$ is the output-sensitive term. 
Since $k \sim O(m^2)$ in the worst case, achieved when $f_1 = - f_0$, we have the following upper bound on $\lvert S_F \rvert$
\begin{equation}\label{eq:sf_m2}
	\lvert S_F \rvert \approx O(m^2)
\end{equation}   
 This bound is exhibited in the grayscale image data example in section~\ref{sec:motivation}: each 9x9 patch contains $(81, 208, 128)$ simplices of dimensions $(0, 1, 2)$ yielding $m=417$ simplices in total which, as the homotopy is simulated, generate $\approx 70,\!000$ transpositions. The upper bound size on $\lvert S_F \rvert$ is thus $\binom{417}{2} \approx 87,\!000$, which nearly occurs in this case, since the last filtration is approximately the reverse of the filtration at the beginning.   
 This quadratic scaling can induce a number of issues in practical implementations of vineyards.
 Indeed, although many transpositions require no column reductions in practice, detecting the existence of non-zero entries at specific positions in the sparse matrix takes $O(\log(m))$ time per transposition.

The ordered set of inversions $(i,j) \in S_F$ can be interpreted as permutations to apply to the decomposition, and hence 
can be thought of as a \emph{schedule} with respect to $F$. 
If our goal is to decrease the size of $\lvert S_F \rvert$, one option is to \emph{coarsen} $S_F$ to a new schedule $\widetilde{S}_F$ by collapsing contiguous sequences of adjacent transpositions to moves, via the map:
\begin{equation}\label{eq:tr_to_mv}
	 (i, i+1)(i+1, i+2)\cdots(j-1, j) \mapsto (j, i+1, \cdots, j-1, i)  \quad \text{ if } i < j
\end{equation}
We expect $\widetilde{S}_F$ to be cheaper to execute than $S_F$, with the tradeoff  that less diagrams are produced.
We make this precise next:
\begin{proposition}\label{prop:factor2}
	Let $R = D V$ be a valid decomposition of size $(m \times m)$, let $C$ denote the number of $O(m)$ operations required to execute schedule $S_F$, and similarly let $L$ be the number of $O(m)$ operations required to execute a coarsened schedule $\widetilde{S}_F$ of $S_F$. We have the following relationship between $C$ and $L$:
	$$ \frac{C}{2} \leq L \leq C$$ 
\end{proposition}
\begin{proof} \normalsize
Consider executing the vineyards algorithm on a given schedule $S_F$.
	%Consider Algorithms~\ref{alg:tr} and~\ref{alg:mr}, where Algorithm~\ref{alg:tr} is used to execute a given schedule $S$ and Algorithm~\ref{alg:mr} is used to execute its coarsened version $\widetilde{S}$. If maps are created providing $O(1)$ access to the low entries needed by lines (5) and (12), and the non-zero entries in each column are sorted according to some fixed order providing $O(\log(m))$ time for lines (5), (11), (17), and (20), then  
	%If we maintain a map locating the position of the lowest entries in $R$ in $O(1)$ time, 
	The dominant subcomputation for the vineyards algorithm are the column operations---which take $O(m)$ time---since the permutations can be made to the matrix in $O(1)$ time and all other computations take at most $O(\log m)$ time. In the vineyards algorithm, there are at most $2$ column operations for a given adjacent transposition per matrix. 
	Thus, for any contiguous sequence of transpositions $(i,i+1), (i+1,i+2), \ldots (j-1,j)$ we require at most $2(\lvert i - j \rvert)$ column operations in both $R$ and $V$, giving a total of $4(\lvert i - j \rvert)$ column operations.

	Now consider using the MoveRight($i$,$j$) for a single move operation, as a replacement for the sequence of transpositions above. Here, the dominant cost again are the column operations (line 5), since although permutations take $O(\lvert i - j\rvert)$ time we require only $1$ such permutation per movement. 
	If we allocate an extra $O(m)$ memory space for the donor columns $d_\ast$ prior to the movement, notice that the assignment to and from $d_\ast$ (lines (4), and (7) in RestoreRight of MoveRight in Algorithm~\ref{alg:mr}) can be completed in $O(1)$ time by storing the column memory addresses as pointers and swapping their pointers. When $d_\mathrm{low}' < d_\mathrm{low}$, instead of copying $\mathrm{col}_\ast(k)$ to $d_\ast'$, which takes $O(m)$ time, we swap their column pointers in $O(1)$ and proceed by modifying the donor column $d_\ast$ with the necessary column operations. After the operations, $d_\ast$ contains the newly modified column and $\mathrm{col}_\ast(k)$ contains the unmodified donor $d_\ast'$, so we swap them once more in $O(1)$ time. 
Since we require at most $1$ $O(m)$ column operation for each index in $[i: j]$, moving a column from $i$ to $j$ where $i < j$ requires at most $2(\lvert i - j \rvert)$ column operations for both $R$ and $V$. The claimed inequality follows. 
	%Identifying non-zero entries in the matrix occurs once, which takes $O(\lvert i - j \rvert \log(m))$ time.
\end{proof}
In terms of bounds, we have $\lvert \widetilde{S}_F \rvert \leq \lvert S_F \rvert$, and the associated coarsened schedule $\widetilde{S}_F$ can be computed in $O(m)$ time. 
However, the coarsening depends entirely on the choice of $F$ and the upper bound from~\eqref{eq:sf_m2} remains, as its always possible that there are no contiguous subsequences to collapse. If a straight line homotopy is used, the number of collapsible subsequences may depend on the simplices categorization as critical points; transpositions between pairs of simplices which trigger  changes in homology tend to be more expensive. Cohen-Steiner~\cite{cohen2006vines} referred to these as \emph{switches}, see~\cite{edelsbrunner2000topological} for more details.

\subsubsection{The Moves Setting}\label{sec:moves_setting}
As we mentioned before, vineyards maintains $j-i$ valid decompositions while 
performing the sequence $(i,i+1), (i+1,i+2), \ldots, (j-1,j)$ of adjacent transpositions. 
The benefit of move operations is that one need not maintain a valid decomposition during the movement $(i,j)$, only at the end; thus, it is worth investigating whether the vineyard bounds can be improved. 
In the case of vineyards, the bound is clear: if $\kappa = K_\tau(K_0, K_1)$ is the Kendall-$\tau$ distance, then $\Omega(\kappa)$ adjacent transpositions are required to transform $K_0 \mapsto K_1$. 
In contrast, if we allow move operations in any order, we have $O(m)$ as a trivial upper bound: simply move each simplex in $K_0$ into its position in the filtration given by $K_1$ in the order given by the latter. However, it's not immediately clear whether this bound is tight.

To establish this formally, we require a few definitions. Let $S_m$ denote the symmetric group  of   permutations of  $[m]$ under function composition $\circ$. Given two fixed permutations $p, q \in S_m$ and a set $\Sigma \subseteq S_m$,  a common  problem is to find an ordered sequence of permutations $s_1, s_2, \dots, s_d \in \Sigma$ whose composition transforms $p$ into $q$:
	\begin{equation}\label{eq:sorting}
		s_d \circ \cdots \circ s_2 \circ s_1 \circ p = q
	\end{equation}
Common variations of this problem include finding such a sequence of minimal length ($d$) and bounding the length $d$ as a function of $m$. In the latter case, a lower bound on $d$ is referred to as the \emph{distance} between $p$ and $q$ with respect to $\Sigma$. 
A sequence $S = \{s_1, s_2, \dots, s_d\} \subseteq \Sigma \subseteq S_m$ of operations mapping $p \mapsto q$ is sometimes called a \emph{sorting of $p$}. 
When $p, q$ are interpreted as strings, distances defined with respect to a fixed $\Sigma \subseteq S_m$ are commonly referred to as \emph{edit distances}~\cite{labarre2013lower}, which are denoted as $d_\Sigma(\cdot, \cdot)$, and the subsequent elementary operations on the strings are called \emph{edit operations}. 
Note the choice of $\Sigma$ defines the universe of operations and thus the type of edit distance being measured---otherwise if $\Sigma = S_m$, then $d_\Sigma(p, q) = 1$ for any $p\neq  q \in S_m$.

Both the difficulty of obtaining a minimal sorting and the size of $d_\Sigma(p,q)$ vary dramatically with the choice of $\Sigma$. 
For example, while sorting by transpositions and reversals is NP-hard, and sorting by prefix  transpositions is unknown, there are polynomial time algorithms for sorting by block interchanges, exchanges, and prefix exchanges~\cite{labarre2013lower}. Sorting by adjacent transpositions can be achieved in many ways: any sorting algorithm that exchanges two adjacent elements during its execution (e.g. bubble sort, insertion sort, merge sort) yields a sorting of size $K_\tau(p, q)$.

In the context of filtered simplicial filtration, we refer to sortings that repair the decomposition as \emph{schedules}. 
\begin{definition}[Schedule]\label{def:schedule}
Given a pair of filtrations $K_0, K_1$ of equal size $m$ and a fixed bijection $f : K_0 \to K_{1}$, let $f^\ast : [m] \to [m]$ denote its induced bijection on $S_m$. A \emph{schedule}  $\mathcal{S} = \{s_1, s_2, \dots, s_d \}$ is a sequence of permutations mapping $K_0 \mapsto K_1$ that satisfies: 
	\begin{equation}\label{eq:rv_seq}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	where, for each $i \in [d]$, $R_i = D_i V_i$ is a valid decomposition: $R_i$ is \emph{reduced} and $V_i$ is full rank upper-triangular. 
\end{definition}
% Note that one cannot arbitrarily permute a given filtration $K$ and its corresponding decomposition: one must respect the face poset.
% Given a filtered simplicial complex $K$, denote by $R = D V$ the decomposition of its corresponding boundary matrix $\partial_K$ such that $R$ is reduced and $V$ is upper-triangular.
% If one has a pair of filtrations $K, K'$ of equal size $m = \lvert K \rvert = \lvert K' \rvert$ and $R = \partial_{K} V$ has been computed, then it may be advantageous to use the information stored in ($R$, $V$) to reduce the computation of $R' = \partial_{K'} V'$. Our first result makes this more explicit.
%   \begin{proposition}\label{prop:preprocessing_time}
%   Given a decomposition $R = \partial_{K} V$ of a simplicial filtration $K$, algorithm~\ref{alg:schedule} can produce such a sequence $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ of permutations with size $\lvert \mathcal{S} \rvert \sim O(m)$ which induces a sequence of decompositions: 
%   	\begin{equation}\label{eq:rv_seq}
%   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
%   	\end{equation}
%	in $O(m \log \log m)$ time and $O(m)$ memory.

%In the context of performing updates to the $R = D V$ decomposition using moves,  such sequences will form schedules. 
%There are typically many ways to solve (1), and any solution to (2) necessarily solves (1) and (3).

%Many types of edit distances can be described succinctly. 
%Let $\beta(i,j,k,l)$ with $1 \leq i < j \leq k < l \leq m$ denote the permutation that exchanges the two closed intervals given by $i, j-1$ and $k,l-1$:
%\[
%\setlength\arraycolsep{1.5pt}
%\scalefont{0.90}{
%\setcounter{MaxMatrixCols}{20}
%\left (
%\begin{array}{*3c|*4c|*4c|*4c|*{5}c } 
%\cline{4-7} \cline{12-15}  \rule{0pt}{2ex}
%1 & \cdots & i -1 & i & i + 1 & \cdots & j - 1 & j & j +1 & \cdots & k - 1  & k & k+1 & \cdots & l - 1 & l & l + 1 & \cdots & m \\
%\cline{4-7} \cline{12-15} 
%\cline{4-7} \cline{12-15}  \rule{0pt}{2ex}
%1 & \cdots & i -1 & k & k+1 & \cdots & l - 1 & j & j +1 & \cdots & k - 1  & i & i + 1 & \cdots & j - 1 & l & l + 1 & \cdots & m \\
%\cline{4-7} \cline{12-15} 
%\end{array}
%\right ) 
%}
%\]
%Common edit distances found in the literature expressed as specializations of $\beta$ are given below:
%	\item (no restriction on $i,j,k,l$) $\implies$ $\beta$ is called a \emph{block interchange}
%	\item $j = k$ $\implies$ $\beta$ swaps adjacent intervals, called a \emph{transposition}
%	\item $j = i+1$ and $l = k+1$ $\implies$ $\beta$ swaps two not necessarily adjacent elements, called an \emph{exchange}\footnote{In comparative genomics, exchanging two elements is sometimes called a \emph{super short transposition}.}
\noindent Here we consider sorting by moves. Phrased using permutations, a \emph{move operation} $m_{ij}$ that moves $i$ to $j$ in $[m]$, for $i < j$, corresponds to the permutation: 

>>>>

 \begin{enumerate}[labelsep=3pt, topsep=2pt,itemsep=-0.25ex,parsep=1.0ex]\label{inv:decomposition}

@@

Property M2 implies the decomposition is not fully maintained during the execution of \textit{RestoreRight} and \textit{RestoreLeft} below, which starkly contrasts the vineyards algorithm. 
In this way, we interpret move operations as making a tradeoff in granularity: whereas a sequence of adjacent transpositions $(i, i{+}1), (i{+}1, i{+}2), \dots, (j{-}1, j)$ generates $\lvert i - j \rvert $ valid decompositions in vineyards, an equivalent move operation $\mathrm{Move}(i,j)$ generates only one.
Indeed, Property M3 directly follows from this fact, as one may simply move each simplex $\sigma \in K$ into its order in $K'$ via insertion sort. Note that the number of valid decomposition produced by vineyards is bounded above by $O(m^2)$ if each pair of simplices $\sigma_i, \sigma_j \in K$ switches its relative ordering at most once during the interpolation from $K$ to $K'$.

@@

%\subsection{Proposed approach}
Let us begin with a brief overview of the pipeline, outlined in Algorithm~\ref{alg:schedule} below.
As before, we assume as input  a discrete $1$-parameter family of filtrations $\mathcal{K} = (K_1, K_2, \dots, K_n)$ of an abstract simplicial complex $K$ with $\lvert K \rvert =m$, and the goal being to compute their persistence diagrams.
Fix reindexing bijections $\rho_i : K_i \to K_{i+1}$. 
Each $\rho_i$ induces a bijection $\rho_i^\ast : [m] \to [m]$, or equivalently, a permutation of the index set $[m]$. For each pair of filtrations $(K_i, K_{i+1})$, we assign the simplices of $K_i$ labels from index set $[m]$ in filtration order, relabeling $K_{i+1}$ accordingly using $\rho_i$. 
As $p$ is now a strictly increasing sequence, we may compute the longest increasing subsequence $\mathrm{LIS}(q)$ and use this subsequence to recover a longest common subsequence (LCS) between $(K_i, K_{i+1})$, which we denote later with $\mathrm{LCS}(K_i, K_{i+1})$. We pass $q$ and $\mathrm{LIS}(q)$ to our greedy scheduling algorithm, which returns as output an ordered set of move permutations  $\mathcal{S}$ of minimum size, which we call a \emph{schedule}.
Note one need not explicitly store the entire family of filtrations nor the schedules between them---Algorithm~\ref{alg:schedule} may be easily modified to be completely \emph{online}, keeping at most two filtrations and one decomposition in memory at any given time. 
% Not all valid schedules are created equal: some incur a much larger number of column operations, even if they have a minimal number of moves; we address this issue in Section~\ref{sec:schedule_cost}.
% Given the set of schedules for each pair $(K_i, K_{i+1})$, we proceed to computing an $R=DV$ decomposition for the first filtration $K_1$, and then execute the moves stored in $\mathcal{S}$ in sequence. During execution, we obtain a valid decomposition $R = DV$ for each filtration $K_1, K_2, \dots, K_n$ from which we extract persistence diagrams along the way.

@@

\subsubsection{Continuous setting}\label{sec:continuous_setting}
In the continuous time case (i.e., vineyards), we are given a homotopy $F : K \times [0,1] \to \mathbb{R}$ interpolating between  a pair of filtrations $K, K'$ from some underlying complex $S$, with simplices ordered according to $f = F(\cdot, 0)$ and $f' = F(\cdot, 1)$, respectively. 
As the choice of homotopy $F$ determines the number of transpositions that transform $K$ into $K$ via $F$, bounding this number requires assumptions on $F$. 
If we assume the curves $\tau \mapsto F(\cdot, \tau)$ are in general position in $[0,1]\times \mathbb{R}$ and that each pair of curves cross  at most once, then the  functions $\tau\mapsto F(\cdot, \tau)$ belong to a class of $x$-monotone curves called \emph{pseudo-segments}.
%, i.e. parameterized curves whose behavior with respect to a certain restricted set of geometric predicates is invariant. 
This family includes the straight-line homotopy $F(\sigma, \tau) = (1 - \tau) f(\sigma) + \tau f'(\sigma)$,  studied in the original vineyards paper~\cite{cohen2006vines}. 
Detecting all $k$ intersections of $m$ pseudo-segments is a well-studied problem in computational geometry that can be optimally solved in output-sensitive $O(m \log m + k)$ time by several algorithms~\cite{boissonnat2000efficient}, where $k$ is the output-sensitive term.

If one has two distinct filtrations $K,K'$ both derived from the same underlying complex $S$ of size $m = \lvert S \rvert$, observe the simplexwise representations of both $K$ and $K'$ can be thought of as elements of the symmetric group $S_m$.
Moreover, as the application of permutations to $K$ implies additional elementary column operations to the underlying matrix decomposition, we think of the action of any set of such permutations on the initial decomposition $R=DV$ of $K$ as a \emph{schedule} with respect to the pair $K,K'$. We formalize this with a definition.
\begin{definition}[Schedule]\label{def:schedule}
Given a pair of filtrations $K, K'$ from an underlying complex $S$ and a fixed bijection $\rho : K \to K'$, let $\rho^\ast : [m] \to [m]$ denote its induced bijection on $S_m$. Given a decomposition $R = DV$ of $K$, a \emph{schedule}  $\mathcal{S} = (s_1, s_2, \dots, s_d )$ is a sequence of permutations mapping satisfying: 
	\begin{equation}\label{eq:rv_seq}
   		R = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R'
   	\end{equation}
   	where, for each $i \in [d]$, $R_i = D_i V_i$ is a valid decomposition respecting invariants~\ref{inv:decomposition}, and $R'$ is a valid decomposition of $K'$.
\end{definition}
\noindent Let $F : K \times [0,1] \to \mathbb{R}$ denote a homotopy of $x$-monotone curves. Then  $F$ decomposes into an ordered set of adjacent transpositions $S_F$ given by pairs of simplices in $K$ that change in their relative ordering between $K$ and $K'$:
\begin{align*}\label{eq:sf_schedule}
	S_F &= \{ (i,i+1) \, \mid \, f(\sigma_i) < f(\sigma_{i+1}), \;  f'(\sigma_i) > f'(\sigma_{i+1}), \; i \in [m] \} \\
	&= \{ s_1, s_2, \dots, s_k \} 
\end{align*}
Moreover, let $p = [m]$ and let $q = \mathrm{Im}\,\rho^\ast$, where $\rho^\ast : [m] \to [m]$ is the one-to-one correspondence induced by the reindexing bijection $\rho: K \to K'$. Then $S_F$ is given by the inversions between $p$ and $q$:
\[S_F = 
\mathrm{Inv}(p, q) = \{ \, (i,i+1) \, \mid \, p(i) < p(i+1), \; q(i) > q(i+1), \; i \in [m]\, \}
\]
The cardinality of $  S_F $ is exactly the \emph{Kendall}-$\tau$ distance~\cite{diaconis1977spearman} between $p$ and $q$: 
\begin{equation}\label{eq:kendall_dist}
	K_\tau(p, q) = \lvert \, \mathrm{Inv}(p, q) \, \rvert
\end{equation}
This implies the $\lvert S_F \rvert \sim O(m^2)$ in the worst case, achieved when $f = - f'$, which is exhibited in the grayscale image data example in section~\ref{sec:motivation}: each 9x9 patch contains $(81, 208, 128)$ simplices of dimensions $(0, 1, 2)$ yielding $m=417$ simplices in total. As the homotopy is simulated, observe $\approx 70,\!000$ transpositions are generated. The upper bound size on $\lvert S_F \rvert$ is thus $\binom{417}{2} \approx 87,\!000$, which nearly occurs in this case, since the last filtration is nearly the reverse of the initial one.

If our goal is to decrease the size of $\lvert S_F \rvert$, one option is to \emph{coarsen} $S_F$ to a new schedule $\widetilde{S}_F$ by collapsing contiguous sequences of adjacent transpositions to moves, via the map:
\begin{equation}\label{eq:tr_to_mv}
	 (i, i+1)(i+1, i+2)\cdots(j-1, j) \mapsto (j, i+1, \cdots, j-1, i)  \quad \text{ if } i < j
\end{equation}
We expect $\widetilde{S}_F$ to be cheaper to execute than $S_F$, with the tradeoff  that less diagrams are produced.
We make this precise next:
\begin{proposition}\label{prop:factor2}
	Let $K$ denote a filtration of size $\lvert K \rvert = m$ with valid decomposition $R = D V$, $T$ the maximum number of $O(m)$ operations required to execute schedule $S_F$ using vineyards, and $M$ be the maximum number of $O(m)$ operations required to execute the coarsened schedule $\widetilde{S}_F$ with moves. Then:
	$$ M \leq {\textstyle \frac{1}{2} \, T} $$ 
\end{proposition}
\begin{proof} \normalsize
First, consider executing the vineyards algorithm on a given schedule $S_F$. As there are at most $2$ column operations, any contiguous sequence of transpositions $(i,i+1), (i+1,i+2), \ldots (j-1,j)$ induces at most $2(\lvert i - j \rvert)$ column operations in both $R$ and $V$, giving a total of $4(\lvert i - j \rvert)$ column operations.

Now consider a single MoveRight($i$,$j$) outlined in Algorithm~\ref{alg:mr}. Here, the dominant cost again are the column operations (line 5). 
Though we need an extra $O(m)$ storage allocation for the donor columns $d_\ast$ prior to the movement, notice that the assignment to and from $d_\ast$ (lines (4), and (7) in RestoreRight of MoveRight in Algorithm~\ref{alg:mr}) can be completed in $O(1)$ time via a pointer swapping argument. That is, when $d_\mathrm{low}' < d_\mathrm{low}$, instead of copying $\mathrm{col}_\ast(k)$ to $d_\ast'$---which takes $O(m)$ time---we swap their column pointers in $O(1)$ before proceeding with the necessary column operations. After the movement, $d_\ast$ contains the newly modified column and $\mathrm{col}_\ast(k)$ contains the unmodified donor $d_\ast'$, so we swap them once more in $O(1)$ time. 
Since we require at most one $O(m)$ column operation for each index in $[i, j]$, moving a column from $i$ to $j$ where $i < j$ requires at most $2(\lvert i - j \rvert)$ column operations for both $R$ and $V$. The claimed inequality follows. 
%since although permutations take $O(\lvert i - j\rvert)$ time we require only $1$ such permutation per movement. 
	%Identifying non-zero entries in the matrix occurs once, which takes $O(\lvert i - j \rvert \log(m))$ time.
\end{proof}
\noindent 
In terms of size bounds, though clearly $\lvert \widetilde{S}_F \rvert \leq \lvert S_F \rvert$ and the associated coarsened $\widetilde{S}_F$ requires just $O(m)$ time to compute, the coarsening depends entirely on the initial choice of $F$---thus the quadratic upper bound remains, as its always possible that there are no contiguous subsequences to collapse. 
As mentioned in~\ref{sec:vineyards}, this quadratic scaling induces a number of issues in the practical implementations of the vineyards algorithm.
% Indeed, although many transpositions require no column reductions in practice, detecting the existence of non-zero entries at specific positions in the sparse matrix takes $O(\log(m))$ time per transposition.

\subsection{Discrete setting}

%If a straight line homotopy is used, the number of collapsible subsequences may depend on the simplices categorization as critical points; transpositions between pairs of simplices which trigger  changes in homology tend to be more expensive. Cohen-Steiner~\cite{cohen2006vines} referred to these as \emph{switches}, see~\cite{edelsbrunner2000topological} for more details.

%\subsubsection{The Moves Setting}\label{sec:moves_setting}
%As we mentioned before, vineyards maintains $j-i$ valid decompositions while 
%performing the sequence $(i,i+1), (i+1,i+2), \ldots, (j-1,j)$ of adjacent transpositions. 
%The benefit of move operations is that one need not maintain a valid decomposition during the movement $(i,j)$, only at the end; thus, it is worth investigating whether the vineyard bounds can be improved. 
%In the case of vineyards, the bound is clear: if $\kappa = K_\tau(K_0, K_1)$ is the Kendall-$\tau$ distance, then $\Omega(\kappa)$ adjacent transpositions are required to transform $K_0 \mapsto K_1$.

Contrasting the continuous-time setting, if we discard the use of a homotopy interpolation and allow move operations in any order, we have $\widetilde{S} \sim O(m)$ trivially: simply move each simplex in $K$ into its position in the filtration given by $K'$ in the order given by the latter. However, it's not immediately clear whether this bound is tight, or whether this strategy of generating schedules---which we call \emph{naive strategy}---is performant.  
%Indeed, arbitrarily permuting a given filtration $K$: one must respect the face poset.

To establish this formally, we require a few definitions. Given two fixed permutations $p, q \in S_m$ and a set $\Sigma \subseteq S_m$,  a common  problem is to find a sequence of permutations $s_1, s_2, \dots, s_d \in \Sigma$ whose composition satisfies:
	\begin{equation}\label{eq:sorting}
		s_d \circ \cdots \circ s_2 \circ s_1 \circ p = q
	\end{equation}
Common variations of this problem include finding such a sequence of minimal length ($d$) and bounding the length $d$ as a function of $m$. In the latter case, a lower bound on $d$ is referred to as the \emph{distance} between $p$ and $q$ with respect to $\Sigma$. 
A sequence $S = \{s_1, s_2, \dots, s_d\} \subseteq \Sigma \subseteq S_m$ of operations mapping $p \mapsto q$ is sometimes called a \emph{sorting of $p$}. 
When $p, q$ are interpreted as strings, distances defined with respect to a fixed $\Sigma \subseteq S_m$ are commonly referred to as \emph{edit distances}~\cite{labarre2013lower}, which are denoted as $d_\Sigma(\cdot, \cdot)$, and the subsequent elementary operations on the strings are called \emph{edit operations}. 
Note the choice of $\Sigma$ defines the universe of operations and thus the type of edit distance being measured---otherwise if $\Sigma = S_m$, then $d_\Sigma(p, q) = 1$ for any $p\neq  q \in S_m$.

Perhaps surprisingly, small changes to set of allowable edit operation $\Sigma$ can dramatically affect both the size of $d_\Sigma(p,q)$ and the difficulty of obtaining a minimal sorting. 
For example, while sorting by transpositions and reversals is NP-hard and sorting by prefix  transpositions is unknown, there are polynomial time algorithms for sorting by block interchanges, exchanges, and prefix exchanges~\cite{labarre2013lower}. Sorting by adjacent transpositions can be achieved in many ways: any sorting algorithm that exchanges two adjacent elements during its execution (e.g. bubble sort, insertion sort) yields a sorting of size $K_\tau(p, q)$.

Here we consider sorting by moves. Phrased using permutations, a \emph{move operation} $m_{ij}$ that moves $i$ to $j$ in $[m]$, for $i < j$, corresponds to the permutation: 
